### 2015-12-05 paper/4-performance.tex
cleaned up R script + put phold figure back in paper
Add memory usage phold and offer explaination for increased memory usage conservative.

\subsubsection{PHold}
PHold \cite{PHOLD} is a parallel oriented benchmark, sequential runtime is measured only as a baseline. A model selects the destination of a message at runtime using a state-saved RNG, which takes up a non-trivial amount of runtime.

@@

\subsection{Parallel Simulation}
%TODO explain that we will now compare different synchronization protocols
%TODO compare different synchronization protocols with adevs' conservative, showing that we are faster (or at least, not that much slower) than adevs
\subsubsection{Devstone}
The flattened models are allocated to kernels by giving each kernel a distinct section of the chain, resulting in a low ratio of inter-kernel to intra-kernel messages. For optimistic, this can cause more reverts since the kernels will start to drift faster as the model-count increases. Furthermore, optimistic is quite sensitive to an increase in kernels, since the delay before a revert propagates increases. Of note as well is the warm-up time this benchmark requires, for $n=d\times w$ models, it takes $\textup{timeadvance}()*(n-1)$ transitions to activate the last model in the chain. For parallel this can be reduced to $\textup{timeadvance}()*(n\frac{kernels-1}{kernels} -1)$ before the last kernel becomes active.\\
\begin{figure}[ltbh]
	\includegraphics[width=.5\textwidth]{fig/fig2.eps}
	\label{fig2.eps}
\end{figure}
\subsubsection{PHold}
%Phold
In Phold allocation is specified in the benchmark itself. Each kernel manages a single node with a constant set of sub-nodes. The parameter R (here 10) determines the percentage of remote destination models.\\
The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing conservative. Lookahead is $\epsilon$, so conservative spends most of its time crawling in steps of $\epsilon$. Since the dependency graph between kernels is a complete graph, this is not a simulation that scales in our implementation. For N kernels, each kernel has to query the null-time of N-1 kernels, resulting in O($N^2$) polling behaviour. In a non-cyclic simulation with a non-trivial lookahead (e.g. devstone), that choice however does pay off. Adevs's lesser performance is due in part to their lookahead management, which after profiling shows to spend a non-trivial amount of time in exception handling code. 
Optimistic suffers little from the above problems, however due to the high interconnectivity a cascading revert is still possible. With the percentage of remotes = 100 phold reflects interconnect in behaviour, which is why the R value is not dimensioned here. A revert is very expensive in PHold due to our usage of C++11's random nr generators. The cost of a revert is dominated by the recalculation of destination models, not in allocating/deallocating states/messages. Again, this could be significantly reduced using lazy-cancellation. Once a revert happens the drift between kernels increases fast, increasing the likelihood of more reverts. Despite all this, optimistic can for low R values quickly exploit uncertainty that slows down conservative in this benchmark.
\subsubsection{Interconnect}
In Interconnect
the set of atomic models form a complete graph (w.r.t. connections), each model broadcasts messages to the entire set. \\
Allocation cannot avoid cycles, the resulting dependency graph between kernels remains a complete graph. The runtime dependency graph is almost immediately equal to the static graph.
Conservative still faces the same issues as in PHold, with the key difference that for a fixed time advance lookahead is equal to the timespan between transitions. The scaling issue is identical as with PHold.\\
In this benchmark optimistic runs very quickly out of memory. With c kernels and N atomic models, a single revert undoing k transitions will lead to $(N-1)\times k$ messages that need to be recreated, plus $(N-1) \frac{c-1}{c} \times k$ anti-messages that need to be send.

@@

%TODO (keep explaining as much as space permits, as this is the evaluation of the core contribution)
\subsection{Memory Usage}
%TODO compare memory usage with adevs for some benchmarks
\subsubsection{Platform and tools}
Both dxex and adevs use tcmalloc as memory allocator. Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (malloc/free/sbrk/mmap/...). Tcmalloc will only gradually release memory back to the OS, whereas our pools will not do so at all. If memory has been allocated once, it is from a performance point of view better to keep that memory in the pool. This is one reason why memory utilization is best measured by peak allocation. Profiling is done using Valgrind's massif tool. \cite{Nethercote:2007:VFH:1273442.1250746}
Platform used for memory profiling has an i5-3317U Intel cpu and 8GiB RAM with a page size of 4,096KiB, running Fedora 22 (kernel 4.2.6).\\
\subsubsection{Measure}
Adevs passes messages by value (in container by reference), we pass by pointer. The runtime effects of this choice are already demonstrated in the interconnect benchmark, so in this section, we measure memory usage in number of allocated pages combining text, stack and heap memory for the program profiled. For the OS and/or user this is the actual memory footprint of the application. It is important to note that, especially in the case of optimistic, not all this memory is always in use by the kernels. During simulation, the pools will generally not return deallocated memory to the OS, but keep it for later reuse.
\subsubsection{Results}
\paragraph*{Devstone}
\begin{table}[lhtb]
	\centering
	\caption{Devstone 40x40 t5e5, unit MiB, 4 kernels (if parallel)}
	\label{dtone_mem}
	\begin{tabular}{| l | l | l | l | l |}
		\hline
		Adevs & AdevsCon &DX &DXCon &DXOpt\\ \hline
		44 & 70 & 42 & 75 & 363  \\ \hline
	\end{tabular}
\end{table}
Since conservative passes messages by pointer, it needs a GVT/LBTS implementation to organise garbage collection. This inevitable delay explains the higher memory usage compared to adevs.\\
%TODO next sentence needs rewriting --done?
Optimistic's TimeWarp requires state/event saving, and its GVT algorithm is more complex (with a resultant higher latency) than the LBTS calculation in conservative. 
Additionally, the differences in LP virtual times are far larger compared to conservative time synchronization. All these factors explain the heavier memory usage. Devstone (flattened) is allocated in a chain. Leafs in the dependency graph will therefore do a lot of unnecessary simulation before having a revert, leading to severe memory pressure. Unlike conservative and sequential execution, memory usage in optimistic varies greatly depending on scheduling of kernel threads and drifting between kernels. 
\paragraph*{Interconnect}
In section 4.2 the parallel performance of this benchmark is further explained. Interconnect highlights dxex's lower performance due to message allocation overhead.
\begin{table}[lhtb]
	\centering
	\caption{Interconnect w 10 t5e5, unit MiB, 2 kernels (if parallel)}
	\label{iconn_mem}
	\begin{tabular}{| l | l | l | l | l |}
		\hline
		Adevs & AdevsCon &DX &DXCon & DXOpt\\ \hline
		39 & 39 & 35 & 43 & 164 \\ \hline
	\end{tabular}
\end{table}

>>>>

\subsubsection{PHold}
PHold \cite{PHOLD} is a parallel oriented benchmark, sequential runtime is measured only as a baseline. A model selects the destination of a message at runtime using a state-saved RNG, which takes up a non-trivial amount of runtime.\\
\begin{figure}[h]
	\includegraphics[width=.5\textwidth]{fig/fig4.eps}
	\label{fig4.eps}
\end{figure}

@@

\subsection{Parallel Simulation}
%TODO explain that we will now compare different synchronization protocols
%TODO compare different synchronization protocols with adevs' conservative, showing that we are faster (or at least, not that much slower) than adevs
\subsubsection{Devstone}
The flattened models are allocated to kernels by giving each kernel a distinct section of the chain, resulting in a low ratio of inter-kernel to intra-kernel messages. For optimistic, this can cause more reverts since the kernels will start to drift faster as the model-count increases. Furthermore, optimistic is quite sensitive to an increase in kernels, since the delay before a revert propagates increases. Of note as well is the warm-up time this benchmark requires, for $n=d\times w$ models, it takes $\textup{timeadvance}()*(n-1)$ transitions to activate the last model in the chain. For parallel this can be reduced to $\textup{timeadvance}()*(n\frac{kernels-1}{kernels} -1)$ before the last kernel becomes active.\\
\begin{figure}[ltbh]
	\includegraphics[width=.5\textwidth]{fig/fig2.eps}
	\label{fig2.eps}
\end{figure}
\subsubsection{PHold}
%Phold
In Phold allocation is specified in the benchmark itself. Each kernel manages a single node with a constant set of sub-nodes. The parameter R (here 10) determines the percentage of remote destination models.\\
The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing conservative. Lookahead is $\epsilon$, so conservative spends most of its time crawling in steps of $\epsilon$. Since the dependency graph between kernels is a complete graph, this is not a simulation that scales in our implementation. For N kernels, each kernel has to query the null-time of N-1 kernels, resulting in O($N^2$) polling behaviour. In a non-cyclic simulation with a non-trivial lookahead (e.g. devstone), that choice however does pay off. Adevs's lesser performance is due in part to their lookahead management, which after profiling shows to spend a non-trivial amount of time in exception handling code. 
Optimistic suffers little from the above problems, however due to the high interconnectivity a cascading revert is still possible. With the percentage of remotes = 100 phold reflects interconnect in behaviour, which is why the R value is not dimensioned here. A revert is very expensive in PHold due to our usage of C++11's random nr generators. The cost of a revert is dominated by the recalculation of destination models, not in allocating/deallocating states/messages. Again, this could be significantly reduced using lazy-cancellation. Once a revert happens the drift between kernels increases fast, increasing the likelihood of more reverts. Despite all this, optimistic can for low R values quickly exploit uncertainty that slows down conservative in this benchmark.

\subsubsection{Interconnect}
In Interconnect
the set of atomic models form a complete graph (w.r.t. connections), each model broadcasts messages to the entire set. \\
Allocation cannot avoid cycles, the resulting dependency graph between kernels remains a complete graph. The runtime dependency graph is almost immediately equal to the static graph.
Conservative still faces the same issues as in PHold, with the key difference that for a fixed time advance lookahead is equal to the timespan between transitions. The scaling issue is identical as with PHold.\\
In this benchmark optimistic runs very quickly out of memory. With c kernels and N atomic models, a single revert undoing k transitions will lead to $(N-1)\times k$ messages that need to be recreated, plus $(N-1) \frac{c-1}{c} \times k$ anti-messages that need to be send.

@@

%TODO (keep explaining as much as space permits, as this is the evaluation of the core contribution)
\subsection{Memory Usage}
%TODO compare memory usage with adevs for some benchmarks
\subsubsection{Platform and tools}
Both dxex and adevs use tcmalloc as memory allocator. Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (malloc/free/sbrk/mmap/...). Tcmalloc will only gradually release memory back to the OS, whereas our pools will not do so at all. If memory has been allocated once, it is from a performance point of view better to keep that memory in the pool. This is one reason why memory utilization is best measured by peak allocation. Profiling is done using Valgrind's massif tool. \cite{Nethercote:2007:VFH:1273442.1250746}
Platform used for memory profiling has an i5-3317U Intel cpu and 8GiB RAM with a page size of 4,096KiB, running Fedora 22 (kernel 4.2.6).\\
\subsubsection{Measure}
Adevs passes messages by value (in container by reference), we pass by pointer. The runtime effects of this choice are already demonstrated in the interconnect benchmark, so in this section, we measure memory usage in number of allocated pages combining text, stack and heap memory for the program profiled. For the OS and/or user this is the actual memory footprint of the application. It is important to note that, especially in the case of optimistic, not all this memory is always in use by the kernels. During simulation, the pools will generally not return deallocated memory to the OS, but keep it for later reuse.
\subsubsection{Results}
\paragraph*{Devstone}
\begin{table}[htb]
	\centering
	\caption{Devstone 40x40 t5e5, unit MiB, 4 kernels (if parallel)}
	\label{dtone_mem}
	\begin{tabular}{| l | l | l | l | l |}
		\hline
		Adevs & AdevsCon &DX &DXCon &DXOpt\\ \hline
		44 & 70 & 42 & 75 & 363  \\ \hline
	\end{tabular}
\end{table}
Since conservative passes messages by pointer, it needs a GVT/LBTS implementation to organise garbage collection. This inevitable delay explains the higher memory usage compared to adevs.\\
%TODO next sentence needs rewriting --done?
Optimistic's TimeWarp requires state/event saving, and its GVT algorithm is more complex (with a resulting higher latency) than the LBTS calculation in conservative. 
Additionally, the differences in LP virtual times are far larger compared to conservative time synchronization. All these factors explain the heavier memory usage. Devstone (flattened) is allocated in a chain. Leafs in the dependency graph will therefore do a lot of unnecessary simulation before having a revert, leading to severe memory pressure. Unlike conservative and sequential execution, memory usage in optimistic varies greatly depending on scheduling of kernel threads and drifting between kernels. 
\paragraph*{PHold}
\begin{table}[htb]
	\centering
	\caption{Phold n 4 s 16 t1e6 r 10, unit MiB, 4 kernels (if parallel)}
	\label{dtone_mem}
	\begin{tabular}{| l | l | l | l | l |}
		\hline
		Adevs & AdevsCon &DX &DXCon &DXOpt\\ \hline
		40 & x & 37 & 61 & 682  \\ \hline
	\end{tabular}
\end{table}
With only 10\% of all messages being inter-kernel, we expect conservative to have memory consumption near that of the single threaded implementation, since intra-kernel messages are reclaimable after each round. The counterintuitive high memory usage can be explained by conservative's stalled round behaviour which occurs whenever a kernel cannot advance (eit $==$ time). In such a round messages are sent out but the kernel does not execute any transitions until it has received all input from all influencing kernels. \\
With lookahead = $\epsilon$ this then leads to a high frequency of polling on shared null-times, which are used to determine lbts and thus garbage collection.
The lbts calculation will not wait until a new value is found, since this can create unwanted contention with the simulation. The resulting longer time intervals within which no new lbts is found delay memory deallocation.
Adevs' conservative fails to complete the benchmark under valgrind even with a significantly reduced load.\\
Optimistic exhibits the expected high memory usage.
\paragraph*{Interconnect}
In section 4.2 the parallel performance of this benchmark is further explained. Interconnect highlights dxex's lower performance due to message allocation overhead. 
\begin{table}[lhtb]
	\centering
	\caption{Interconnect w 10 t5e5, unit MiB, 2 kernels (if parallel)}
	\label{iconn_mem}
	\begin{tabular}{| l | l | l | l | l |}
		\hline
		Adevs & AdevsCon &DX &DXCon & DXOpt\\ \hline
		39 & 39 & 35 & 43 & 164 \\ \hline
	\end{tabular}
\end{table}
